---
title: "hff_stats"
author: "Tom Su"
date: "2025-08-12"
output: html_document
---


# INITIAL SETUP

This block clears the workspace and imports additional packages needed for the
analysis:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list = ls())   # empty the workspace
library(ggplot2)  # for plotting
library(emmeans)  # generating emmeans
library(lme4)     # for linear mixed effects regression modeling
library(lmerTest) # for post-hoc
#library(kenward-roger) # for post-hoc
library(ggthemes) # `yutannihilation.github.io/
                  # allYourFigureAreBelongToUs/ggthemes/`
library(car)     # qqplots
library(interactions) #interactions visualisation
library(DHARMa) # outlier visualisation
citation("DHARMa")

library(bestNormalize)
library(simr)
```


# STUDY DESIGN AND HYPOTHESES

This study examines the effect of continuous pulsatile electrical stimulation on
somatosensory-fibre response rates. Data was collected in LabChart using single-
fibre teased-fibre recordings and processed in Python prior to import for
statistical analysis in this document.

## Aim 1: Effect of latency on rate of decay and asymptote over 5 minutes

* dependent variables: binned response rate // (fitted inverse tau and
asymptote)
* independent variables: time (log transform) // latency
* data: binned values by time (log transform in Python or R?) // fitted values
vs latency
~EXPORTED~

## Aim 2: Effects of frequency and amplitude on somatosensory-fibre response
## rates

* dependent variables: response rate
* independent variables: frequency/amplitude, latency(?), sum of current
delivered
* data: short-duration trial average response rate for each trial over 3s
* group by unit id
~EXPORTED~

* also look at recovery in this section - what actually needs to be exported?
what variables are we interested in? -- for recovery, defined as rolling bin (5 samples) at 4/6 or 6/6

## Aim 3: Interactions between mechanical and electrical stimulation

* dependent variables: response rate
* independent variables: interleaved vs regular, latency(?)
* data: 9:1 trial average response rate for each trial over 3s
~EXPORTED~
* data: 50 Hz interleaved after long duration - also mech recov here?
  * what to export? rolling bins?
* group by unit id

## Aim 4: Amplitude/latency changes

* dependent variables: amplitude/latency
* independent variables: frequency, stim amp, latency?, sum of current delivered
* data: 3s trials and 5 minute trials (not binned)


## Potential Supplementary data: Cell type classification effect on amplitude,
## latency, asymptote

* dependent variables: amplitude, latency, asymptote | response rate
* independent variables:  Unit classification (LTFA, LTSA, HTSA)


# VARIABLES AND DATA HEADINGS

## DATA FILES

Filenames are given as the aim number and then a description of the information

### Aim 1

Associated files:

2_binned-values (0.1s bins) = response rate bins with a 0.1 s bin width
2_binned-values = response rate bins with a 1 s bin width
2_fitted-values = exponential decay fitted values long duration data
TBD = long duration 1 second responses



### Aim 3

## VARIABLE NAMES

    animal_id = Unique animal identifier
    unit_id = Animal ID_cell number
    test_id = concatenated ID in format trial type_frequency_amplitude
    trial_id = full concatenated ID in format unit ID_trial type_frequency_amplitude multipler_trial number
    sex = M or F
    unit_type = Qualitative mechanical classification
    conduction_velocity = calculated conduction velocity in m/s 

    A = constant for long duration data exponential fit (k1)
    k = 1/τ for long duration data
    b = constant for long duration data exponential fit (k2)
    c = asymptote for long duration data
    rmse = root mean squared error of the fit

    time = Bin centre
    response_probability = average probability of firing at a specific bin timing
    log_probability = Natural log of response probability

# DATA IMPORT

## Aim 1

```{r}

fittedvalues <- read.csv("datafiles/1_fitted-values.csv") # this makes a dataframe (df). You call the columns as follows: df$column_name.

binvalues <- read.csv("datafiles/1_binned-values.csv") # this makes a dataframe (df). You call the columns as follows: df$column_name.

```

## Aim 2

```{r}
frequencyvalues <- read.csv("datafiles/2_frequency-values.csv") # this makes a dataframe (df). You call the columns as follows: df$column_name.

amplitudevalues <- read.csv("datafiles/2_amplitude-values.csv") # this makes a dataframe (df). You call the columns as follows: df$column_name.
```

## Aim 3

```{r}
nineonevalues <- read.csv("datafiles/3_nineone-values.csv") # this makes a dataframe (df). You call the columns as follows: df$column_name.
```

## Aim 4
```{r}
longspikevals  <- read.csv("datafiles/4_amplitude-latency-fitted-values.csv")
spikevals <- read.csv("datafiles/4_amplitude-latency-sweeps-values.csv")

#first 30 stims

spikevals30 <- read.csv("datafiles/4_amplitude-latency-sweeps-values_first30.csv")


# this makes a dataframe (df). You call the columns as follows: df$column_name.
```
# STATISTICAL ANALYSIS

## Aim 1

```{r, include=FALSE}

#old fit checking RMSE

#rmse <- fittedvalues

#rmse$sqrmse <- (fittedvalues$rmse)^2

#models <- c("single", "double")

#modelfactors <- factor(models)

#rmse$rmse_desc <- as.factor(rmse$rmse_desc)

#mod3 <- lmer(sqrmse ~ rmse_desc + (1|unit_id),  data = rmse)
#plot(residuals(mod3) ~ predict(mod3, re.form = NA))
#x<-resid(mod3)
#qqp(x) 

#summary (mod3)


```

Fit the models and check assumptions
```{r}

#factor code if needed; not for this study
#data$subject <- as.factor(data$subject)

#based of QQ plots, logged data for better normal distribution

#asymptote model

#fittedvalues <- subset(fittedvalues, fittedvalues$c < 1)


fittedvalues$logc <- log(fittedvalues$c)
mod0 <- lmer(logc ~ conduction_velocity + (1|animal_id),  data = fittedvalues)
x<-resid(mod0)
qqp(x)
plot(residuals(mod0) ~ predict(mod0, re.form = NA))

#time constant model

#fittedvalues <- subset(fittedvalues, fittedvalues$unit_id != "hff11_1")

#fittedvalues$logk1 <- log(fittedvalues$k1)

mod1 <- lmer(k ~ conduction_velocity + (1|animal_id),  data = fittedvalues)
x<-resid(mod1)
qqp(x)
plot(residuals(mod1) ~ predict(mod1, re.form = NA))


#also set up a test for unit type but given how low the Ns are, did not think this was appropriate to test or discuss further. Leaving here for completeness.
#fittedvalues$logc <- log(fittedvalues$c)
#mod100 <- lmer(logc ~ unit_type + (1|animal_id),  data = fittedvalues)
#x<-resid(mod100)
#qqp(x)
#plot(residuals(mod100) ~ predict(mod100, re.form = NA))

#summary (mod1b)
#anova (mod1b)
#emmeans((mod1b), pairwise ~ unit_type)

#mod1b <- lmer(k ~ unit_type + (1|animal_id),  data = fittedvalues)
#x<-resid(mod1b)
#qqp(x)
#plot(residuals(mod1b) ~ predict(mod1b, re.form = NA))

#emm1 <- emmeans(mod1b, ~ unit_type)
#emm1
#summary (mod1b)
#anova (mod1b)
#emmeans((mod1b), pairwise ~ unit_type)

```

Residuals look good, running comparisons on the emmeans

```{r}
#c model
emm0 <- emmeans(mod0, ~ conduction_velocity)
summary (mod0) 
anova (mod0)
#k model

emm1 <- emmeans(mod1, ~ conduction_velocity)
summary (mod1) 
anova (mod1)
#emm2 <- emmeans(mod1b, ~ conduction_velocity)
#summary (mod1b) 

emm100 <- emmeans(mod100, ~ unit_type)
summary (mod100) 
anova (mod100)

```

Plottinginverse tau and asymptote values
```{r}

# asymptote values (c)
ggplot(fittedvalues, aes(conduction_velocity, c)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)



#inverse tau (k)
ggplot(fittedvalues, aes(conduction_velocity, k)) + 
#  xlim(0, 40) +
#  ylim(0, 0.5) +
  geom_point()


```

## Aim 2

Analyse whether there is a relationship between response probability and frequency, amplitude at 3 seconds

```{r}

frequencyvalues$logresponse_probability_all <- log(frequencyvalues$response_probability_all)

#model warns about rescaling, but rescaled frequency does not change P values, so leaving alone

#data$subject <- as.factor(data$subject)

#QQ plots show high skew so log transforming this dataset

mod4 <- lmer(logresponse_probability_all ~ frequency * conduction_velocity + (1|animal_id:unit_id),  data = frequencyvalues)
plot(residuals(mod4) ~ predict(mod4, re.form = NA))
x<-resid(mod4)
qqp(x) 

summary (mod4)
anova (mod4)

#suspicious outlier in the residuals, running DHARMa to confirm
#simres <- simulateResiduals(mod4)
#plot(simres)
#testOutliers(simres)
#nothing detected so leaving it in and moving on.

#testing unit type here as well - nothing found
mod74 <- lmer(logresponse_probability_all ~ frequency * unit_type + (1|animal_id:unit_id),  data = frequencyvalues)
plot(residuals(mod74) ~ predict(mod4, re.form = NA))
x<-resid(mod74)
qqp(x) 

summary (mod74)
anova (mod74)


#interact_plot(mod4, pred = frequency, modx = conduction_velocity, plot.points = TRUE)

jn <- johnson_neyman(mod4, pred = conduction_velocity, modx = frequency)
jndata <- jn$cbands
write.csv(jndata, "johnson_neyman_output.csv", row.names = FALSE)

jn

#plot <- subset(frequencyvalues, frequencyvalues$frequency == "200")

plot <- frequencyvalues
ggplot(plot, aes(frequency, response_probability_all)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)

frequencyvalues$logresponse_probability_initial <- log(frequencyvalues$response_probability_initial)

mod5 <- lmer(logresponse_probability_initial ~ frequency * conduction_velocity + (1|animal_id:unit_id),  data = frequencyvalues)
plot(residuals(mod5) ~ predict(mod5, re.form = NA))
x<-resid(mod5)
qqp(x) 
summary (mod5)
anova (mod5)


ggplot(frequencyvalues, aes(frequency, response_probability_all)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)


ggplot(frequencyvalues, aes(frequency, response_probability_all)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)

#log transform does not improve skew for amplitude so not logging this set

mod64 <- lmer(response_probability_all ~ amplitude * conduction_velocity + (1|animal_id:unit_id),  data = amplitudevalues)
plot(residuals(mod64) ~ predict(mod64, re.form = NA))
x<-resid(mod64)
qqp(x) 
summary (mod64)
anova (mod64)

ggplot(amplitudevalues, aes(amplitude, response_probability_all)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)


```
Power Analysis

```{r}

#for the long duration

#power_simr0 <- powerSim(fit = mod0, test = fixed("conduction_velocity"), nsim = 1000)
#power_simr0

#for the short duration
#power_simr1 <- powerSim(fit = mod4, test = fixed("frequency"), nsim = 1000)
#power_simr1


```

## Aim 3

```{r}

#nineonevalues$test <- as.factor(nineonevalues$test)
#nineonevalues$stimulus <- as.factor(nineonevalues$stimulus)

nineonevalues$logr <- nineonevalues$response_probability
mod6 <- lmer(response_probability ~ test * stimulus + (1|animal_id:unit_id),  data = nineonevalues)
plot(residuals(mod6) ~ predict(mod6, re.form = NA))
x<-resid(mod6)
qqp(x) 

summary (mod6)

anova (mod6)
emmeans(mod6, pairwise ~test|stimulus)

```

## Aim 4

Analyse whether changes in spike timing are significant

```{r}

#yj_obj <- yeojohnson(spikevals$cv_normal_mean)
#spikevals$normcv <- yj_obj$x.t

#longspikevals  <- read.csv("datafiles/6_amplitude-latency-fitted-values.csv")
#spikevals <- read.csv("datafiles/6_amplitude-latency-sweeps-values.csv")

#longspikevals <- subset(longspikevals, longspikevals$unit_id != "hff11_1")
spikevals <- subset(spikevals, spikevals$unit_id != "hff11_1") #one outlier removed from these datasets after testing outliers using simulated residuals - 11_1, which saw a visual sudden jump in amplitude/timing halfway through the recording set.
#spikevals <- read.csv("datafiles/4_amplitude-latency-sweeps-values.csv")


# first long duration cvs

normcv <- longspikevals$latency_actual_steady_cv - longspikevals$initial_conduction_velocity
t.test(normcv, mu = 0)

normlat <- longspikevals$latency_actual_c - longspikevals$initial_latency
t.test(normlat, mu = 0)



mod7 <- lmer(latency_normal_steady_cv ~ initial_conduction_velocity + (1|animal_id),  data = longspikevals)
plot(residuals(mod7) ~ predict(mod7, re.form = NA))
x<-resid(mod7)
qqp(x) 

summary(mod7)
anova (mod7)


ggplot(longspikevals, aes(initial_conduction_velocity, latency_normal_steady_cv)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)

#freq and amp sweep cvs


freqspikevals <- subset(spikevals, spikevals$test == "frequency")

mod8 <- lmer(cv_normal_mean ~ test_frequency * initial_conduction_velocity + (1|animal_id:unit_id),  data = freqspikevals)
plot(residuals(mod8) ~ predict(mod8, re.form = NA))
x<-resid(mod8)
qqp(x)



summary(mod8)
anova (mod8)

simres <- simulateResiduals(mod8)
plot(simres)
testOutliers(simres)


ggplot(freqspikevals, aes(test_frequency, cv_normal_mean)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)

freqspikevalsunit <- subset(freqspikevals, freqspikevals$test_frequency == "150")

ggplot(freqspikevalsunit, aes(unit_type, cv_normal_mean)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)


ampspikevals <- subset(spikevals, spikevals$test == "amplitude")

mod9 <- lmer(cv_normal_mean ~ test_amplitude * initial_conduction_velocity + (1|animal_id:unit_id),  data = ampspikevals)
plot(residuals(mod9) ~ predict(mod9, re.form = NA))
x<-resid(mod9)
qqp(x) 

summary(mod9)
anova (mod9)



ggplot(ampspikevals, aes(test_amplitude, cv_normal_mean)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)


# this makes a dataframe (df). You call the columns as follows: df$column_name.

```
Analyse whether changes in spike amplitude are significant


```{r}

#longspikevals <- subset(longspikevals, longspikevals$unit_id != "hff11_1")
spikevals <- subset(spikevals, spikevals$unit_id != "hff11_1") #one outlier removed from these datasets after testing outliers using simulated residuals - 11_1, which saw a visual and sudden jump in spike timing halfway through the recording.

#spikevals <- read.csv("datafiles/4_amplitude-latency-sweeps-values.csv")

# first long duration cvs

#test slope and intercept

normamp <- longspikevals$amplitude_actual_intercept - longspikevals$initial_amplitude
t.test(normamp, mu = 0)
t.test(longspikevals$amplitude_actual_slope, mu = 0)

mod10 <- lmer(amplitude_normal_mean ~ initial_conduction_velocity + (1|animal_id),  data = longspikevals)
plot(residuals(mod10) ~ predict(mod10, re.form = NA))
x<-resid(mod10)
qqp(x) 

summary(mod10)
anova (mod10)


ggplot(longspikevals, aes(initial_conduction_velocity, amplitude_normal_mean)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)

#freq and amp sweep cvs

freqspikevals <- subset(spikevals, spikevals$test == "frequency")

mod11 <- lmer(amplitude_normal_mean ~ test_frequency * initial_conduction_velocity + (1|animal_id:unit_id),  data = freqspikevals)
plot(residuals(mod11) ~ predict(mod11, re.form = NA))
x<-resid(mod11)
qqp(x) 

summary(mod11)
anova (mod11)

simres <- simulateResiduals(mod11)
plot(simres)
testOutliers(simres)


ggplot(freqspikevals, aes(test_frequency, amplitude_normal_mean)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)

ampspikevals <- subset(spikevals, spikevals$test == "amplitude")

mod12 <- lmer(amplitude_normal_mean ~ test_amplitude * initial_conduction_velocity + (1|animal_id:unit_id),  data = ampspikevals)
plot(residuals(mod12) ~ predict(mod12, re.form = NA))
x<-resid(mod12)
qqp(x) 

summary(mod12)
anova (mod12)


ggplot(ampspikevals, aes(test_amplitude, amplitude_normal_mean)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)


# this makes a dataframe (df). You call the columns as follows: df$column_name.

```
Analyse whether changes in spike timing are significant in first 30

```{r}

spikevals30 <- subset(spikevals30, spikevals30$unit_id != "hff11_1") 

#freq and amp sweep cvs


freqspikevals30 <- subset(spikevals30, spikevals30$test == "frequency")

mod8 <- lmer(cv_normal_mean ~ test_frequency * initial_conduction_velocity + (1|animal_id:unit_id),  data = freqspikevals30)
plot(residuals(mod8) ~ predict(mod8, re.form = NA))
x<-resid(mod8)
qqp(x)

summary(mod8)
anova (mod8)



ggplot(freqspikevals30, aes(test_frequency, cv_normal_mean)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)

freqspikevals30unit <- subset(freqspikevals30, freqspikevals30$test_frequency == "150")

ggplot(freqspikevals30unit, aes(unit_type, cv_normal_mean)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)

ampspikevals30 <- subset(spikevals30, spikevals30$test == "amplitude")

mod9 <- lmer(cv_normal_mean ~ test_amplitude * initial_conduction_velocity + (1|animal_id:unit_id),  data = ampspikevals30)
plot(residuals(mod9) ~ predict(mod9, re.form = NA))
x<-resid(mod9)
qqp(x) 

summary(mod9)
anova (mod9)

ggplot(ampspikevals30, aes(test_amplitude, cv_normal_mean)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)



```

Analyse whether changes in spike amplitude are significant in first 30

```{r}


#longspikevals30 <- subset(longspikevals30, longspikevals30$unit_id != "hff11_1")
spikevals30 <- subset(spikevals30, spikevals30$unit_id != "hff11_1") #one outlier removed from these datasets after testing outliers using simulated residuals - 11_1, which saw a visual and sudden jump in spike timing halfway through the recording.

#freq and amp sweep cvs

freqspikevals30 <- subset(spikevals30, spikevals30$test == "frequency")

mod11 <- lmer(amplitude_normal_mean ~ test_frequency * initial_conduction_velocity + (1|animal_id:unit_id),  data = freqspikevals30)
plot(residuals(mod11) ~ predict(mod11, re.form = NA))
x<-resid(mod11)
qqp(x) 

summary(mod11)
anova (mod11)


ggplot(freqspikevals30, aes(test_frequency, amplitude_normal_mean)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)

ampspikevals30 <- subset(spikevals30, spikevals30$test == "amplitude")

mod12 <- lmer(amplitude_normal_mean ~ test_amplitude * initial_conduction_velocity + (1|animal_id:unit_id),  data = ampspikevals30)
plot(residuals(mod12) ~ predict(mod12, re.form = NA))
x<-resid(mod12)
qqp(x) 

summary(mod12)
anova (mod12)
emmeans(mod12, pairwise ~ test_amplitude|initial_conduction_velocity)


#simres <- simulateResiduals(mod9)
#plot(simres)
#testOutliers(simres)


ggplot(ampspikevals30, aes(test_amplitude, amplitude_normal_mean)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)


jn <- johnson_neyman(mod12, pred = initial_conduction_velocity, modx = test_amplitude)
jndata <- jn$cbands
write.csv(jndata, "johnson_neyman_output2.csv", row.names = FALSE)

jn

```



```{r}


ggplot(fittedvalues, aes(unit_type, c)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)


freqspikevalsunit <- subset(freqspikevals, freqspikevals$test_frequency == "100")



ggplot(freqspikevalsunit, aes(unit_type, cv_normal_mean)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)

frequencyvaluesunit <- subset(frequencyvalues, frequencyvalues$frequency == "100")

ggplot(frequencyvaluesunit, aes(unit_type, response_probability_all)) + 
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = FALSE)

write.csv(fittedvalues, "fittedvalues.csv", row.names = FALSE, quote = FALSE)

```
